---
title: "R Notebook"
output: html_notebook
---
```{r}
knitr::opts_knit$set(root.dir = normalizePath("/Users/alexisjimenez/Desktop")) 
```

```{r}
library(dplyr)
library(ggplot2)

#load ancestry (ADMIXTURE Q file)
ancestry_data <- read.table("/Users/alexisjimenez/Desktop/pruned_data.3.Q", header=FALSE)

# Load fam file 
fam_data <- read.table("/Users/alexisjimenez/Desktop/pruned_data.fam", header=FALSE)
colnames(fam_data) <- c("Family_ID", "Sample", "Paternal_ID", "Maternal_ID", "Sex", "Phenotype")

# Add family ID f
ancestry_data <- cbind(Family_ID = fam_data$Family_ID, ancestry_data)

# Rename ancestry columns to match Admixture groups
colnames(ancestry_data) <- c("Family_ID", "South_Asian", "African", "European")

```
edit the fam file 
```{r}

library(stringr)

# get rid of the array data in ID and only leave numeric 
fam_data$Sample_ID <- str_extract(fam_data$Sample, "(?<=_)[0-9]+(?=_)")
head(fam_data)

# Reorder columns so that the column names match between file 
fam_data <- fam_data[, c("Family_ID", "Sample_ID", setdiff(names(fam_data), c("Sample_ID", "Family_ID")))]

head(fam_data)

```

```{r}
#load miRNAs, I manually added the Family_IDs to a column to make it easier to merge the data 
miRs <- read.csv("/Users/alexisjimenez/Desktop/miRs_with_FamilyID.csv")


```


```{r}
library(dplyr)

# Merge ancestry and  miRs by Family_ID
merged_data <- inner_join(ancestry_data, miRs, by = "Family_ID")

# Merge with fam_data to include any additional participant metadata
merged_data <- inner_join(merged_data, fam_data, by = "Family_ID")

head(merged_data)

#colnames(merged_data)

```

```{r}
# Remove duplicate columns from merging all the files 
merged_data <- merged_data %>%
  select(-c(Family_ID.1, Sample_ID.x, Sample_ID.y))


colnames(merged_data)

```
```{r}
# Get rid of non-numeric columns and only keep miRs
miRNA_data <- merged_data %>%
  select(-c(Family_ID, Sample, PTID, BOX_POSITION, MATERIAL, 
            SONGLAB.BARCODE, DATEACQ, DATE.PURIFIED, PURIFIED.BY, 
            whole.blood.est.vol..ml., NANODROP.DATE, ng.µL, 
            X260.280, X260.230, DNA.STOCK.VOLUME..ML., DNA.YIELD..µg., 
            dna.stock.vol.for.Illumina.array..µL., NOTES, 
            Paternal_ID, Maternal_ID, Sex, Phenotype))

#get rid of group columns 
miRNA_data <- miRNA_data %>% select(-c(Group2, Group3))

#get rid of the admixture columsn
miRNA_data_pca <- miRNA_data %>%
  select(-c(South_Asian, African, European))

# convert everything to numeric 
miRNA_data_pca <- data.frame(lapply(miRNA_data_pca, as.numeric))

# Run PCA on miRNA expression data only
pca_result <- prcomp(miRNA_data_pca, scale. = TRUE)


```

```{r}
# Making a new dataframe with PCA results and ancestry proportions for visualization
pca_components <- data.frame(pca_result$x, miRNA_data %>% select(South_Asian, African, European))

```


```{r}
#making a PCA with the admixture groups allowing for overlap of Admixture groups 
ggplot(pca_components, aes(PC1, PC2, color = as.factor(South_Asian))) +
  geom_point() +
  labs(color = "Admixture Group") +
  xlab("PC1") +
  ylab("PC2") +
  ggtitle("PCA of miRNA Expression by Ancestry Group") +
  theme_minimal()

```
```{r}


```{r}
# Create a new column for dominant ancestry group for no overlap of ancestry groups 
miRNA_data <- miRNA_data %>%
  mutate(Admixture_Group = case_when(
    South_Asian > African & South_Asian > European ~ "South Asian",
    African > South_Asian & African > European ~ "African",
    European > South_Asian & European > African ~ "European",
    TRUE ~ "Admixed" 
  ))

```

```{r}
# Merge PCA results with ancestry group labels
pca_components <- data.frame(pca_result$x, Admixture_Group = miRNA_data$Admixture_Group)

```

```{r}
library(ggplot2)

#create a new PCA plot that assigns partipants to group they fit the best in 
ggplot(pca_components, aes(x = PC1, y = PC2, color = Admixture_Group)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "PCA of miRNA Expression by Admixture Group",
       x = "Principal Component 1",
       y = "Principal Component 2",
       color = "Admixture Group") +
  theme_minimal()

```

```{r}
write.csv(merged_data, file = "/Users/alexisjimenez/Desktop/PRYSM_ADMIXTURE_miRNA.csv", row.names = FALSE)

```

