---
title: "PRS and outcome"
author: "Alexis Jimenez"
date: "2025-04-24"
output: html_document
---
#knit
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("/Users/alexisjimenez/Desktop")) 
```

#load libraries
```{r}
library(tidyverse)
library(vcfR)
library(stringr)
library(missMethods)
```
#laod the data 
```{r}
PRS<- read.csv("/Users/alexisjimenez/Desktop/DPP_HivertPRS_Prelim.csv")
outcome <- read.csv("/Users/alexisjimenez/Desktop/DPP_2yr_DM_outcome.csv")
ID_links<- read.csv("/Users/alexisjimenez/Desktop/DPP_IDs_linked_miRs_and_genes.csv")
```

```{r}
# Clean the X column to keep only the part before the first underscore
PRS$X <- sub("_.*", "", PRS$X)

```

#visualize data
```{r}
head(PRS$X)
head(ID_links$SUBJECT_ID)

```


```{r}
# Make sure both are character type
PRS$X <- as.character(PRS$X)
ID_links$SUBJECT_ID <- as.character(ID_links$SUBJECT_ID)

# merge
merged_dataset <- merge(PRS, ID_links, by.x = "X", by.y = "SUBJECT_ID")

colnames(merged_dataset)
```
```{r}
merged_data$RELEASE_ID <- as.character(merged_data$RELEASE_ID)
outcome$RELEASE_ID <- as.character(outcome$RELEASE_ID)

final_data <- merge(merged_data, outcome, by = "RELEASE_ID")

colnames(final_data)
```

```{r}
library(dplyr)
final_data_genetic <- final_data %>% 
  filter(!is.na(prs_risksum))  # Keep people who have PRS scores

```

```{r}
# Load packages
#install.packages("survminer")
library(dplyr)
library(survival)
library(survminer)

# Filter for people with PRS scores 
final_data_genetic <- final_data %>%
  filter(!is.na(prs_risksum))

# Create PRS quartiles
final_data_genetic$PRS_quartile <- cut(final_data_genetic$prs_risksum,
                                       breaks = quantile(final_data_genetic$prs_risksum, probs = seq(0, 1, 0.25), na.rm = TRUE),
                                       include.lowest = TRUE,
                                       labels = c("Q1", "Q2", "Q3", "Q4"))

# Create survival objects (diabetes progression or not) 
surv_object <- Surv(time = final_data_genetic$DIABT.2, event = final_data_genetic$DIABF.2)

# Cox proportional hazards model by PRS quartile (adjusting for trial arm)
cox_model_quartiles <- coxph(surv_object ~ PRS_quartile + ASSIGN, data = final_data_genetic)
summary(cox_model_quartiles)

# Plotting survival curves by PRS quartile
surv_fit_quartiles <- survfit(surv_object ~ PRS_quartile, data = final_data_genetic)

ggsurvplot(surv_fit_quartiles, data = final_data_genetic,
           risk.table = TRUE,
           pval = TRUE,
           conf.int = TRUE,
           xlab = "Time (years)",
           ylab = "Survival Probability (Diabetes-Free)")

#Collapse PRS into 2 groups: Q1–Q3 = Low, Q4 = High
final_data_genetic$PRS_group <- ifelse(final_data_genetic$PRS_quartile == "Q4", "High", "Low")

# Cox proportional hazards model for the two groups
cox_model_2groups <- coxph(surv_object ~ PRS_group + ASSIGN, data = final_data_genetic)
summary(cox_model_2groups)

# Plot survival curves for 2 groups
surv_fit_2groups <- survfit(surv_object ~ PRS_group, data = final_data_genetic)

ggsurvplot(surv_fit_2groups, data = final_data_genetic,
           risk.table = TRUE,
           pval = TRUE,
           conf.int = TRUE,
           xlab = "Time (years)",
           ylab = "Survival Probability (Diabetes-Free)")


``````
#fixing formatting 
```{r}
ggsurvplot(
  surv_fit_quartiles,
  data = final_data_genetic,
  risk.table = TRUE,
  pval = TRUE,
  conf.int = TRUE,
  palette = "Dark2",
  ggtheme = theme_minimal(),
  xlab = "Time (Years)",
  ylab = "Diabetes-Free Survival Probability",
  legend.title = "PRS Quartile",
  legend.labs = c("Q1", "Q2", "Q3", "Q4"),
  risk.table.height = 0.25
)

#collapsed quartiles
ggsurvplot(
  surv_fit_2groups,
  data = final_data_genetic,
  risk.table = TRUE,
  pval = TRUE,
  conf.int = TRUE,
  palette = "Set1",
  ggtheme = theme_minimal(),
  xlab = "Time (Years)",
  ylab = "Diabetes-Free Survival Probability",
  legend.title = "PRS Group",
  legend.labs = c("Low", "High"),
  risk.table.height = 0.25
)

```

